Evaluation of clustering parameters

### 2024-08-08 analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Add 02-clustering template analysis

---
title: "Evaluation of clustering parameters - SCPCS000170"
author: maud-p
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: kable
params:
  sample_name: ~
  sample_path: ~
  out_rds_path: ~
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE)
```

@@

In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000170) of the Wilms Tumor dataset (SCPCP000006).

@@

### Packages

Load required packages in the following chunk, if needed.
Do not install packages here; only load them with the `library()` function.

```{r packages, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(Seurat)      # to use Seurat workflow

library(tidyr)
library(dplyr)
library(DT)          # for table visualization

library(ggplot2)     # for visualization
library(patchwork)   # for visualization
library(ggplotify)
library(SCpubr)      # for visualization
library(viridis)     # for visualization - colors

library(edgeR)       # For pseudobulk based differential expression (DE) analysis
library(DElegate)    # for pseudobulk based DE analysis and identification of marker genes

library(Azimuth)     # For label transfer

library(msigdbr)     # for gene set enrichment analysis or pathway enrichment
library(clusterProfiler)

library(data.table)
```

#### Some config set-up / threshold

We store in a config list names "cfg" parameters used all along the analysis to filter for p-value, log fold change, percentage of expression, etc.

```{r cfg}
cfg = list()
cfg$padj_thershold = 0.05
cfg$lfc_threshold = 1
cfg$rate1_threshold = 0.5
set.seed(12345)
```

#### Base directories

```{r base paths, eval=TRUE, include=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "2024-07-08", "SCPCP000006")

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
```

#### Input and output files

For this analysis, we worked with the _processed.rds data. 
We builded a Seurat object based on the counts data and re-perform the analysis 
[normalization --> reduction --> clustering] following the Seurat workflow.

We transferred meta.data to keep:

[x] QC data computed by the DataLab

[x] annotation data computed by the DataLab

[x] raw annotation and gene_symbol conversion

```{r paths}
# input
# use download-data.py to download the data

# get all files in the data directory
filelist <- list.files(data_dir, full.names = TRUE)

# select the 40 Wilms tumor single nucleus folders only
filelist <- filelist[! grepl(".tsv", filelist)]

# output
## html report for each of the Wilms tumor sample will be saved in 
path_to_report <- file.path(module_base, "01-clustering")
## some plots from the report will be saved in 
path_to_plot <- file.path(module_base, "/plots")
## after final decision on the clustering strategy, processed RDS files will be saved in 
path_to_result <- file.path(module_base, "/results")
```

```{r paths2}
## to be modified when rendered
i = 3
filedir <- filelist[i] 
sample <- gsub(paste0(data_dir, "/"), "", filedir)

@@

Here we have a sample of favorable histology, untreated. Few thoughts:

- We thus do not expect anaplasia-associated patterns (TP53 as an example) to be major driver here.

- We do not expect high immune infiltration (induced by chemotherapy).

- We do not expect chemotherapy induced DNA damages.

@@

[8] We run label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. 
We plot pca/umap reduction grouping with latest labels. 
We expect it to be the most representative of the cell types in the sample.

@@

# Normalization
options(future.globals.maxSize= 891289600000)
srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)

@@

C9 and C13 might be a specific subtype, immune, endothelium, healthy cells (?)

@@

C1 might be composed of cycling cells, in higher proportion than the other clusters.

@@

Here, we open the table of marker genes marker-sets/CellType_metadata.csv.

@@

  # open the set of reference marker genes

CellType_metadata <- read.table("marker-sets/CellType_metadata.csv", header = TRUE, sep = ",")

@@

```{r fig.width=25, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("malignant")])] , ncol = 9, group.by = "seurat_clusters", legend.position = "none") 
```
Almost all clusters are NCAM1+ (ENSG00000149294) which is a marker of blastema area.

@@

C9 is positive for COL6A3 which is a marker of stroma Wilms tumor cells (also normal stroma)

@@

Almost all clusters are WT1+ (ENSG00000184937).

@@

Almost all clusters are IGF2+ (ENSG00000167244).

@@

C13 is PTPRC+ (ENSG00000081237) and must be a cluster of immune cells. Wilms tumor are generally cold tumor, especially untreated Wilms tumor. According to the clinical data, the present sample is untreated (treatment = upfront resection). We do not expect high percentage of immune cells. The size of the cluster would therefore make sense.

@@

The set of clusters [0-2-3-4-5-6-11-12] is EPCAM+ (ENSG00000119888), which is a marker of epithelial (malignant and normal) kidney cells.

C9 is slightly VWF+ (ENSG00000110799) and might contain few endothelial cells.

@@

```{r marker_pathways, fig.width=12, fig.height=6, out.width='100%'}
# open the set of reference marker genes

Pathway_metadata <- read.csv("marker-sets/Pathway_metadata.csv", header = TRUE, sep = ",")

DT::datatable(Pathway_metadata, caption = ("CellType_metadata"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))
```
#### TP53 pathway

@@

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```
"Normal" cells (immune, endothelial) have a slightly higher TP53 score. I am however not sure how relevant the TP53 score is for Favorable sample as the one we are analysing here. To be tested on an anaplastic sample.

@@

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```
I am not sure how relevant the DNA repair score is for Favorable sample as the one we are analysing here. To be tested on an anaplastic sample.

@@

- PAX8 (C2), associated woth blastema and epithelial tumor cells (https://pubmed.ncbi.nlm.nih.gov/34306127/)
- CD74 and HLA-DRA in C3 suggest the presence of immune cells in C13.

@@

```{r fig.width=25, fig.height=7, out.width='100%'}
tmp_H <- NULL
for(i in unique(srat$seurat_clusters)){
signature = s.markers$feature[s.markers$group1 == i]
ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_hallmarks)
tmp <- NULL

@@

- C9 is enriched for EMT signature (Warning count = 5) which makes sense with the hypothesis that they are stroma cluster.

@@

```{r fig.width=25, fig.height=6, out.width='100%'}

@@

@@

@@

Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, we can say that C13 a cluster of immune cells.

@@

Note to the DataLab : where should I save the fetal kidney reference? 
In S3 like results as it is a quite big object or in the folder marker-sets?

@@

```{r fig.height=6, fig.width=6, message=FALSE, warnings=FALSE}

# access to the azimuth kidney reference 
##### to be updated depending on the location of the reference file 
ref_dir = file.path(module_base, "marker-sets/Azimuth_Compatible_Fetal_full/")

reference <- LoadReference(ref_dir)
# Load the query object for mapping

# Preprocess with SCTransform
# Note, the RunAzimuth change a bit the Seurat object, changing for example the rownames to Symbol. 
# We thus choose to rename the object to keep the 
data <-  RunAzimuth(srat, ref_dir, assay = 'RNA')
rm(reference)
gc()

DT::datatable(table(data$seurat_clusters, data$predicted.compartment), caption = ("table of azimuth compartment annotation per seurat clusters"),               extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))

DT::datatable(table(data$seurat_clusters, data$predicted.celltype), caption = ("table of azimuth annotation per seurat clusters"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))

```{r fig.height=5, fig.width=20, warnings=FALSE}

d1 <- DimPlot(data, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + NoLegend()
d2 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
d3 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.celltype", label = TRUE, repel = TRUE) + NoLegend()

d1 | d2 | d3
```

Looking at the predicted compartment, we are satisfied that the immune compartment is specific to C13. 
Endothelium and stroma compartment are mixed in C9.

It is expected that the fetal__nephron compartment match to both blastema and epithelial cancer and normal kidney.

Looking at the predicted cell types, we observed that the set of cluster [0,1,2,3,4,5,6,7,8,10,11,12] is labeled as cap mesenchyme, an early developmental stage of nephron
This suggest that clusters [0,1,2,3,4,5,6,7,8,10,11,12] are more stem, either primitive epithelium or blastem cancer cells.

```{r fig.height=5, fig.width=20, warnings=FALSE}

f1 <- SCpubr::do_BarPlot(sample = data,
                         group.by = "predicted.compartment",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

f2 <- SCpubr::do_BarPlot(sample = data,
                         group.by = "predicted.celltype",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

f1 | f2

## Summary and conclusion

Taking together, we can conclude:

- C9 = stroma + endothelium (with high confidency)
- C13 = immune cells (with high confidency)

- C2. C7, C13, that form a unit in the umap reduction, are stromal cells, either cancer or normal (with high confidency)
- [0,1,2,3,4,5,6,7,8,10,11,12] are cancer cells, either blastema or primitive epithelium.

@@

[ ] We will adapt this template to be render on all the 40 Wilms tumor samples of the dataset.

[ ] We will save for each sample the rds file

[ ] We will run inferCNV for each sample to decide the malignant/normal status of some stroma and epithelial cluster, and confirm the blastema annotation.

>>>>

---
title: "Evaluation of clustering parameters"
author: maud-p
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: kable
params:
  sample_name: ~
  sample_path: ~
  out_rds_path: ~
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warnings = FALSE,
	error = TRUE
)
```

@@

In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000169) of the Wilms Tumor dataset (SCPCP000006).

@@

In order to assess the clustering quality, we look into some marker genes, pathways enrichment and label transfer.

## Sample metadata

```{r paths2}
## to be modified when rendered
filedir <- filelist[i] 
sample <- gsub(paste0(data_dir, "/"), "", filedir)

@@

@@

[ ] Next step, aim for a next PR: We will label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. 
We plot pca/umap reduction grouping with latest labels. 
We expect it to be the most representative of the cell types in the sample.

@@

# Normalization
options(future.globals.maxSize= 8912896000000)
srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)

@@

We expect up to 5 set of clusters:

- blastema cancer cells
- epithelium cancer and/or normal cells
- stroma cancer and/or normal cells
- immune cells
- endothelial cells

@@

@@

Here, we open the table of marker genes marker-sets/CellType_metadata.csv. 
Note: we do not expect to have a clear and nice pattern of expression for all of the following markers in every tumor. 
This is just ti get a few idea.

@@

@@

```{r fig.width=25, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("malignant")])] , ncol = 9, group.by = "seurat_clusters", legend.position = "none") 
```

##### NCAM1

@@

##### COL6A3

@@

##### IGF2

##### WT1

@@

##### EPCAM

@@

##### PTPRC = CD45

@@

##### VWF, endothelial cells

@@

```{r marker_pathways, fig.width=12, fig.height=6, out.width='100%'}

DT::datatable(Pathway_metadata, caption = ("Pathway_metadata"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

#### TP53 pathway

@@

"Normal" cells (immune, endothelial) have a slightly higher TP53 score.

@@

Note: Chemo-treated samples should have higher DNA-damage scores.

@@

@@

```{r fig.width=25, fig.height=10, out.width='100%'}
tmp_H <- NULL
for(i in unique(srat$seurat_clusters)){
signature = s.markers$feature[s.markers$group1 == i]
ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_hallmarks)
tmp <- NULL

@@

EMT signature should be enriched in stroma cluster.
E2F/proliferation should be enriched in blastema cluster.
MYC(N), TP53 must be enriched in blastema cluster.

@@

```{r fig.width=25, fig.height=10, out.width='100%'}

@@

Here to check of we catch and MIR pattern (DROSHA; DICER1; other?)

@@

The MSigDB C8 gene set is quite relevant for kidney and nephroblastoma annotations. 
Epithelial (cancer and normal) cells should be enriched in mature/adulte kidney pathways while blastema cancer cells will show enrichment of fetal kidney development pathway / cap mesenchyme.

@@

Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, it give the first impression that C12 is a cluster of endothelial cells and C11 a cluster of immune cells.

@@

To be added in a next PR.

@@

@@

-[x] We will adapt this template to be render on all the 40 Wilms tumor samples of the dataset.

-[ ] We will save for each sample the rds file

-[ ] We will run inferCNV for each sample to decide the malignant/normal status of some stroma and epithelial cluster, and confirm the blastema annotation.
